---
title: "Final Group Project - Proposal"
author: "Abdulaziz Aljuaid, Jialong Li, Jack Lyu, Yu Qin, Xun Yang"
date: "2024-03-20"
output: pdf_document
---

```{r, include = TRUE, warning = FALSE, message = FALSE}
library(tidyverse)
library(modelr)
ev_data <- read_csv("Electric_Vehicle_Population_Data.csv")
demo_data <- read.csv("demo.csv")
```
# Dataset
Electric Vehicle Population Data: This dataset shows the Battery Electric Vehicles (BEVs) and Plug-in Hybrid Electric Vehicles (PHEVs) that are currently registered through the Washington State Department of Licensing (DOL).
Washington State of United States Census Bureau: household income in each county and education.
We have extracted the data from the Census Bureau and filtered out the data with the variables that we are interested in to research on into the file 'demo.csv'.

# Questions
* We are trying to find the EV per capita in Washington State. Start from counties first, and then try to simulate through Census Tract.
* How does EV per capita relate to the geographical information of different counties? 
* Can we predict the adoption rate of EVs for counties, and find counties that are slower on the EV adoption but have the capability of adopting EVs so the companies can make more sales in those counties?

# A Preliminary Model
* Y variable is EV per capita by each county who is 18 years and over.

* X variables are median household income in the county and the percentage of the population that obtained a bachelor's degree or higher that is older than 25.

# ggplot
```{r, include = TRUE, warning = FALSE, message = FALSE}
# number of ev per county in WA state
wa_ev_data <- ev_data %>%
  filter(State == "WA") %>%
  group_by(County) %>%
  summarise(Number_of_EV = n())
print(wa_ev_data)
```

```{r, include = TRUE, warning = FALSE, message = FALSE}
merged_data <- merge(wa_ev_data, demo_data, by = "County")
merged_data <- merged_data %>%
  mutate(EV_per_Capita = Number_of_EV / population)
print(merged_data)
```


```{r, include = TRUE, warning = FALSE, message = FALSE}
merged_data$percentage <- as.numeric(gsub("%", "", merged_data$percentage)) / 100
model <- lm(EV_per_Capita ~ income + percentage, data = merged_data)
summary(model)
```
```{r}
library(ggplot2)

ggplot(merged_data, aes(x = income, y = EV_per_Capita)) +
  geom_point(aes(color = County), alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) + 
  labs(x = "Median Household Income", y = "EV per Capita", title = "EV per Capita vs. Median Household Income") +
  theme_minimal()
```
```{r}
ggplot(merged_data, aes(x = percentage, y = EV_per_Capita)) +
  geom_point(aes(color = County), alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Percentage of Population with Bachelor's Degree or Higher", y = "EV per Capita", title = "EV per Capita vs. Bachelor's Degree or Higher") +
  theme_minimal()
```

# Future Analysis or Datasets
We have explored the relationship between the per capita ownership of electric vehicles (EVs) in each county, median household income, and the percentage of the population over 25 with a bachelor's degree or higher. This provides valuable insights into the adoption of electric vehicles across different regions in Washington state. Here are some variables for future analyses or datasets:

Comprehensive Demographic and Economic Factors

1. Geographical information update

* The current goegraphical variable used is county, but with only 39 data entries, fitting model and get ideal outcome is extremely challenging, for future study, we consider to use census tract, a much smaller geographical block and we will have much more data entries in the state of washington and make it better for us to fit models.

2. Fitting with different models

* The model used for this proposal is linear regression modeling and for our future research we consider to use different model applies to the dataset and try to fit the dataset better with other models ohter than simple linear regression

3. Considering other variables

* For our future studies, we want to consider other variables that might have impact on the adoption of EV associated with different geographical areas.

4. Segmentation of data

* We are also considering using segmented demographic information to see how different age group react to EV and the difference of the EV adaption rate with different age groups.
